Vagueness as cost reduction

### 2016-07-27 papers/2015/internal_response_to_reviews/Kees_response_to_reviews.txt

Hi Matt,

As promised, a short summary of our plans regarding our LCP paper. Perhaps you can take a quick look to see if you think this is where we stand.

===========================================================================================================================

Summary of discussion on our LCP submission, 17 May 2012.

Timing: Your thesis has higher priority, but we should try to keep the momentum, so we can resubmit not too long after our thesis submission (in early October?)

Issues from the reviews:

1. We want to keep the focus on vagueness (contra reviewer 3) but (as suggested by reviewer 1) we need to flag at an early stage that vagueness is difficult to isolate for a number of reasons:

(a) the difference between potential/actual vagueness
(b) the difference between being vague and not containing a number
(c) the difference between vagueness and non-specificity (as in Mishra's work for instance)

2. We need to be clearer perhaps about the research question that we're testing. The suggestion from reviewer 2, to state alternative hypotheses before each experiment, might be good.

3. We need to read all the suggested literature: Dehaene 97, Feigenson et al 2004, Dehaene 2008, Sauerland & Stateva 2007. The latter contends that "about x" is not vague. Maybe a quick check of this paper will suffice to dismiss it? (There could be a link with the issue 1(c) above.)

4. Consider whether we want to measure reading times (as suggested by reviewer 1). If so, what method should we use? Also, should this influence the choice of materials (e.g., should we try harder to find expressions of similar length?) It's hard enough to find suitable materials as it is, so we should probably avoid further complications ...

5. In a new version of Experiment 3, we should probably make the gaps between numbers proportional to the size of the numbers.

6. The big issue: finding out how much evaluation procedures contribute to the "cost" of a description. Concretely, reviewer 1 worries that the difference that we describe to presence/absence of Number might actually be caused by the possibility/impossibility of doing the task on the basis of a comparison strategy (i.e., "choose the largest/smallest of the two").

-- The first suggestion made by he reviewer might be useful: he suggests to compare "find the square with many dots" vs. "find the square with more than x dots". Both of these can be solved by a comparison strategy, so they're alike in that regard. Our hypothesis would be that the first one is easier. This sounds fine, but can't we argue that we have done almost precisely this as part of experiment 3, when we compared "few" and "fewest"?

-- One could think of an analogous experiment where none of the two cases can be resolved using comparison. For example, "with about x dots" vs "with x dots". On the other hand, we did exactly this as part of experiment 3 ... The question that this raises is whether perhaps we can address this reviewer's main criticism without doing a new experiment.

-- A possible new version of experiment 3, to address the comparison strategy confound. Ideally we'd look at all 8 combinations of YesNumber/NoNumber, YesVague/NoVague, YesComparison/NoComparison. "YesComparison" means that the task can be solved by a simple comparison (as opposed to laborious counting). Here's an imperfect attempt:

YN, YV, YC: far fewer than 10
YN, YV, NC: about 10  -- I do find this an intriguing contrast. We could do this as a separate experiment (analogous to the suggestion of reviewer 1).

YN, NV, YC: fewer than 10
YN, NV, NC: 10

NN, YV, YC: few
NN, YV, NC: few but not very few (?) -- Wouldn't subjects simply choose the one in the middle? (What if we used more than 3 numbers?)

NN, NV, YC: fewest
NN, NV, NC: fewest except one (?) -- Problem: this is solvable through a slightly more complex kind of comparison. Can we find a better way to fill this slot?

Can we perhaps argue that the NN part of this table is impossible to complete, so the best we can do is an experiment focussing on YN (the first four items in the list)? This strikes me as an approach that might satisfy reviewer 1.

========================================================================================================================

k.

### 2016-07-27 papers/2015/resubmission_drafts/2015_resubmission_draft/vagueness-dec-2015.tex
adding papers

% changing to a version that only reports experiments 2 to 5
\documentclass[doc,apacite]{apa6}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{color}
\usepackage{booktabs}
\usepackage{layout}
\usepackage{apacite}
\usepackage{datetime}

%\usepackage[draft]{changes}
\usepackage[final]{changes}
\setdeletedmarkup{\textcolor{red}{\sout{#1}}}
%\usepackage[right=2 in]{geometry}
\setlength{\marginparwidth}{72pt}

\title{Vagueness as cost reduction}
\date{\today}
\authornote{\today ~\currenttime}
\shorttitle{Vagueness as cost reduction}
\twoauthors{Matt Green}{Kees van Deemter}
\twoaffiliations{Computing Science\\ University of Aberdeen}{Computing Science\\ University of Aberdeen}

% KvD: Throughout the paper I've tried to use "instruction format" consistently, instead of sometimes "instruction format", sometimes "quantifier format".

\abstract{Much of everyday language is vague, yet standard game-theoretic models make it difficult to see how vague expressions can have benefits over crisp ones in co-operative situations. We report a series of experiments that aims to separate the utility of vagueness from the utility of other factors that come together with vagueness. We argue that the results support a view of vagueness where the benefits that vague terms exert are due to other influences that vagueness brings with it rather than to \deleted{influences of} vagueness itself.}

@@

Linguists, philosophers of language, and more recently game theorists, have asked why natural languages contain so many vague expressions, which are used so frequently \cite{Lipman:2000fk, lipmanvague}. By introducing borderline cases, these expressions create potential misunderstandings, thereby creating ``a worldwide several-thousand year efficiency loss'' \cite<>[p.~1]{lipmanvague}. Lipman explains the point by means of a scenario in which a speaker describes a person to a hearer, who needs to identify that person in the arrivals hall of of an airport. Lipman argues that, in such a scenario, a precise description of the person's height (e.g., ``The person's height is 187.96 cm'') would be more useful than a vague one (``The person is tall''). Lipman uses this scenario to explain why standard game theory models of communication \cite<e.g.,>{Crawford:1982lr} predict that, under certain conditions, a crisp act of communication will always have more utility than a vague act \deleted{of communication} that communicates the same state of affairs. The relevant conditions are, broadly speaking, that both interlocutors know all the relevant facts (e.g., both know the person's height precisely) and that the setting for the communication is co-operative. These conditions exclude deliberate deception, and rhetorical situations like political debate and advertising where the intention is to persuade the interlocutor to adopt some point of view, where the persuasion might not be intended to be in the addressee's best interest.

Lipman \replaced{argued}{observed} that such the efficiency loss resulting from vague expressions would be unlikely to have arisen unless there are advantages as well as disadvantages associated with vague expressions. Lipman asked, essentially, what these advantages might be, and how they might find a place in a game-theoretical explanation. In this paper, we focus on the first part of Lipman's question.

Several tentative answers to Lipman's question have been offered  \cite<see>{van2009utility, van2010vagueness}. Prominent among these answers is the idea that vague expressions are somehow easier to process, by a speaker and/or a hearer, than expressions that are not vague (i.e., crisp) \cite<e.g.,>[]{lipmanvague,De-Jaegher:2003lr,vanrooij2003lr}. For example, \citeA<>[p.\ 11]{lipmanvague} writes: ``For the listener, information which is too specific may require more effort to analyze''. We shall refer to this characterisation of the utility of vague language as the \emph{cost reduction} hypothesis. The idea of vagueness as cost reduction can take various shapes~\deleted{\protect\cite<>[]{van2009utility}}, but the basic hypothesis is that it is easier for people to think in terms of loosely defined categories (such as ``quite a few'', or ``many'') than in terms of crisply defined ones (such as ``thirteen'', or ``237''). This predicts that whenever a vague expression can perform the same communicative task as a precise one, it is rational to choose the vague expression. The corollary for comprehension is that vague expressions should be understood more readily than precise expressions.

Questions concerning optimal language use have many practical applications. Natural Language Generation\footnote{NLG systems take data or formulas as input, and transform them into natural language outputs \cite<>[]{reiter2000building}. The process parallels language production in humans.}  (henceforth NLG) systems must make decisions between different formulations of the same information. For example, if a man's height is 6 foot 2 inches, this could be expressed as ``187.96 metres'', ``6 foot 2'', or ``tall'', among other ways, and the NLG system must decide between these.  The problem is particularly relevant for NLG systems that take numbers as input, as many do. In the context of an NLG system faced with a practical decision of this kind, Lipman's question becomes ``Under what circumstances should vague terms be produced?''  Relevant applications include weather forecasting on the basis of numerical weather data such as temperature and wind speed \cite{goldberg1994using,turner2006generating}, and medical decision support on the basis of clinical measurement such as oxygen saturation, heart rhythm, etc. \cite{Hripcsak01032009, hunter2008summarising, portet2009automatic}. At present, such NLG systems \replaced{are often forced to make}{make} decisions concerning the level of precision in the utterances that they generate (e.g., ``the temperature will be in the high twenties tomorrow'') on the basis of little more than intuition. A better understanding of the benefits of different precision levels for readers would allow these systems to become more useful.

@@

In a \added{series of} studies of behaviour modification, \citeA<>{Mishra01042011} manipulated the presentation format of information about quantities \added{in the domains of mental acuity, physical strength, and weight loss}. \added{In the weight loss study, participants were told that the study was designed to test the validity of a new (actually fictitious) health index, the HHI (Holistic Health Index). They were told that an ideal HHI score lies in the range of 45 to 55. In a longitudinal study, participants submitted their height, weight, hydration level, gender, and age to a computer each week. Participants were told that two algorithms would be used to compute their HHI, and that it was possible that the two algorithms might give different values initially, but would converge over the course of the study to a single value. They were also told that if the two algorithms did give different values, then the true score lay between the two values. In one condition, which the authors called the precise condition, the two algorithms gave the same score. In the other condition, which the authors called the vague condition, one algorithm added 3\% to the score while the other algorithm subtracted 3\% from the score, yielding a range of values whose midpoint was the same as the two values given in the precise condition.

One group of participants was given HHI scores in the ideal range: for this group their weight loss did not differ depending on whether they were given vague or precise HHI values. However for the other group, who were given HHI scores outside the ideal range, their weight loss was significantly greater if they were given vague HHI scores than if they were given precise HHI scores. The authors explain the improvement in the vague condition for this group as resulting from the participants' freedom to think of themselves a positioned on one end of the range - the end closest to the ideal HHI scores.} 
%
This ``illusion of proximity'' \cite<>[p.~4]{Mishra01042011} to the goal is argued to allow participants to generate positive expectancies that lead to behaviours that improve performance. In contrast, in the precise conditions, participants did not have this freedom of interpretation, and could not distort the information to bring about the beneficial \emph{illusion of proximity}. These results are interesting, and of obvious potential practical importance. We note, however, that information presented as an exact range of values does not conform with the standard definition of vagueness \cite{keefe1996vagueness, EgreKlinedinst}, since an exact range does not admit borderline cases. In the terminology of \citeA{Hobbs85granularity}, the difference between a range and a single midpoint value is a difference of \emph{granularity}. Furthermore, the experiments of \citeA{Mishra01042011} did not explore benefits in terms of processing cost, but in terms of long-term behaviour change.

Similar issues arise from the work of \citeA{peters2009bringing}. The authors carried out a series of studies where participants were required to rate hospitals based on various sources of information about quality of care. There was a between-subjects manipulation based on numeracy. The format of the information was manipulated within subjects: either numbers only were presented, or both numbers and evaluative categories were presented (e.g., \emph{Poor}, \emph{Fair}, \emph{Good}, \emph{Excellent}, with \added{crisp} visual boundary lines between the categories). Results showed that, for low-numeracy participants, the presence of evaluative categories resulted in a diminished influence of an irrelevant affective state on the ratings. For all participants, the presence of evaluative categories resulted in better decisions and in a greater use of the most important and reliable types of information, such as survival rates. 
%Participants' decisions were evaluated with respect to a gold-standard set of interpretations derived from experts. 
% KvD The last sentence is worth checking.

It is, however, questionable whether the ``evaluative categories" manipulation in this study can be considered a manipulation of vagueness. Certainly, terms like \emph{Fair} admit the possibility of borderline cases. However, \replaced{given that the boundaries between the categories were marked crisply, and that therefore the categories mapped crisply to numerical values} {when the visual boundary lines are taken into account, which map the terms to exact ranges}, it becomes doubtful whether any borderline cases could be conceived to arise in fact. For example, \emph{Fair} was mapped to 60\% -- 70\% for the variable \emph{percentage of heart attack patients given recommended treatment (ACE inhibitor)}. {\deleted\footnote{This distinction between the potential for vagueness, and the realisation of vagueness in the context of a particular stimulus, is something we will return to when discussing our own experiments.}} Accordingly, rather than the vagueness of categories such as \emph{Poor}, Peters et al. emphasise the evaluative content inherent in these categories, and \replaced{the affective potential of the evaluative content rather than the vagueness of the terms like {\em Fair}}{their affective potential}.\\[2ex]
%
The experiments reported in the present paper put the cost reduction hypothesis to the test. The question that we are trying to answer is whether vague expressions are processed more easily by readers than crisp ones. Like Lipman, we focus on situations where numerical information is used in order to identify a referent. Reference, in other words, will be the linguistic task on which we focus, partly because of the interest that this topic has recently drawn from the NLG community.  In focussing on benefits for the hearer, we will leave aside the question of audience design, leaving this for later research.

\added{In using references to quantities to test the cost reduction hypothesis we are only testing one aspect of vagueness in a particular context. This limits the applicability of our results. It also has the advantage that we can explore the costs and benefits of vagueness more thoroughly in that context. Since one prevalent view of vagueness is that a vague expression is never preferable to a crisp equivalent, a demonstration of a benefit for vagueness in any context would advance the discussion.}

\added{In our experiments we present participants with an instruction to select a referent from among alternatives, using vague and crisp referring expressions to identify the referent. In this context, speed and accuracy of selecting the referent are the key dimensions on which benefits or costs for vagueness should be apparent. We used dot arrays as the possible referents of the instructions, and we varied how the instructions identified a particular dot array, using vague and crisp versions of the instructions. For example, the participant could be instructed `Choose the square with many dots' in the vague condition, or `Choose the square with 20 dots' in the crisp condition.}

\deleted{Game theory and the cost reduction hypothesis both make claims about vagueness in terms of its effects on people engaged in communicative linguistic acts. For game theory, when vagueness influences communication, the term \emph{utility} is used to capture this influence: vagueness is said to have less utility than crispness. The cost reduction hypothesis notes that we use vague language frequently, and assumes that the reason for the high frequency of use is that vagueness brings about an \emph{advantage}, for producer or comprehender. }

\deleted{In this paper, we set out to measure these effects so that we might have some empirical basis for preferring one account over the other. An obstacle to this effort is that whereas we can obtain, in the laboratory setting, several quantitative measures of the effects of language, neither the game theory account nor the cost reduction account specifies a metric in which \emph{utility} or \emph{advantage} should be measured. Therefore it is for us as experimenters to suggest a way to capture and measure the effects of vague and crisp expressions such that they can meaningfully be compared. This requires us to choose, and motivate the choice of, both a task that brings about some measurable behaviour (or alternatively, some measurable physiological state) that is consequent upon whether the language used to bring it about was vague or crisp; and also a measure of the behaviour that is sufficiently fine-grained to distinguish the vague case from the crisp case.}

\deleted{
A good candidate for the task comes from the \emph{forced choice} paradigm in which a participant is required to make a choice quickly among alternatives, indicating the choice by a physical response such as a button-press. These alternatives can be presented using vague and crisp language in such a way that the participant's choice is at least partly dependent upon whether the presentation was vague or crisp: indeed, in the case where other factors are held constant, it is plausible, and often assumed, that differences in the response are consequent to a great extent upon the mode of presentation. The button-press behaviour offers two metrics which are commonly taken to index cognitive load: response time and response accuracy, where fast accurate responses indicate low cognitive load, and slow or inaccurate responses indicate high cognitive load. If vague presentations lead to lower cognitive load than crisp presentations, or vice versa, (where load is indexed by response time and accuracy), it is plausible to argue that this constitutes an advantage, or greater utility, of one presentation mode over the other.}

\deleted{
In the experiments that we present here, we induced people to make a particular choice among alternatives by indicating one alternative using a referring expression.  In the particular situation that we used, participants were presented with a number of objects on screen that constituted the alternatives.  These objects were squares containing a number of dots, where the number in each was different. The referring expression referred to one square by indicating the number of dots it contained, and the participant was required to indicate that square by pressing the appropriate button on the keyboard. The referring expression offered a way to manipulate whether the square was indicated with a crisp or a vague expression of quantity, while holding other factors relatively constant. For example, the participant could be instructed ``Choose the square with many dots'' in the vague condition, or ``Choose the square with 20 dots'' in the crisp condition.}

% Inserting introduction material that mentions the other factors that go along with vagueness in this setup. This is in response to reviewer comments to mention the confounds earlier - MG

In practice manipulating vagueness independently of other factors in these forced choice tasks with numeric expressions is very difficult.
%
\deleted{For example when comparing few/many against small numbers like 2,3,4 (subitizable numbers) it is expected that the specialised routines that the brain has for processing small numbers will come into play: these could overwhelm the influence of vagueness. Experiment One addresses the role of vagueness for subitizable and not-subitizable numbers.} 
%
For example when comparing few/many against \deleted{non-subitizable} numbers like 10, 20, although few/many are vague and 10,20 are crisp, they also differ according to numerical and verbal format. This difference between numerical and verbal format could explain any differences that are found between the vague and crisp conditions for these instructions. \deleted{Experiment Two demonstrates that there are differences for these instructions that can be interpreted in line with the cost reduction hypothesis, while acknowledging that numerical / verbal format could be the underlying driver of the effect.} It is possible to create numerical and verbal versions of both vague and crisp instructions in order to test whether numerical/verbal format drives \replaced{any differences observed between}{the processing advantage for} few/many versus 10/20. \deleted{demonstrated in Experiment Two.} \added{For example, the numerical crisp conditions (10, 20) have numerical vague versions (about 10, about 20); and the verbal vague conditions (few, many) have verbal crisp versions (the fewest, the most).}

\deleted{
Experiment Three takes the existing numerical crisp (10/20) and verbal vague (few, many) conditions and adds instructions in  numerical vague (about 10 / about 20), and verbal crisp (the fewest / the most) expressions. We increase the number of squares to three for this experiment, so that expressions like few and many have a borderline case.
}

Another non-vagueness  difference between few/many and 10/20 is that the process of identifying the square (the selection algorithm for making the decision) might differ. For example, in the vague condition, identifying the square with few dots might be done by a comparison algorithm, identifying that one square is less numerous than the others without establishing the numerosity of either square; whereas, in the crisp condition, establishing that one of the squares has about 10 dots would require an estimate of the numerosity of the squares (a matching algorithm, matching the number mentioned in the instruction to estimates of the numerosity of the squares). Thus if participants respond more quickly when instructed to choose the square with few dots, this could be due to vagueness or to comparison being easier than matching.
%
%\deleted{There is some evidence that the size of the array can influence the selection algorithm used to identify the referent of an instruction. \protect\citeA{Feigenson2004307} discusss two systems for representing number. They write that ``large arrays [\ldots] activate a system for representing sets and comparing their approximate cardinal values'' whereas``small arrays [\ldots] activate a system for representing and tracking numerically distinct individuals, which allows for computations [\ldots] of the number of individuals in the array'' \protect\cite[p.~311]{Feigenson2004307}.}
% KvDTentatively deleted this.

\added{ Experiments 4 and 5 take up this issue and the numeric/verbal format issue, by comparing comparison and matching versions of both vague and crisp instructions, when number-use is held constant: numbers are always mentioned in Experiment 4, while numbers are avoided throughout Experiment 5.
% KvD: I've tentatively reinstated this, because I think we cannot simply state the problem without announcing our approach to it. If you know a better way to do this then that's fine. 
}

%\section{\deleted{Experiment 1}}
% \deleted{
%In our first experiment we set out to compare responses to vague and crisp instructions in a forced choice task using response time and accuracy to measure the responses, and a variety of combinations of numbers. We aimed to identify cases where a difference was elicited such that these cases could form the focus of  subsequent experiments.
%}
%\subsection{\deleted{Method}}
%
%\subsubsection{\deleted{Participants}}
%\deleted{
%Twenty-five students and staff from the University of Aberdeen participated in the study in return for a cash payment of five pounds. Their median age was 23, ranging from 18 to 40. All participants self-reported fluency in English, and had normal (or corrected to normal) vision. Participants were recruited by advertising for participants on a university message board and a university mailing list, offering five pounds to volunteers. 
%}
%
%\subsubsection{\deleted{Apparatus}}\deleted{
%A MacBook Pro laptop computer with a 13 inch screen presented the stimuli to the participants. Stimuli were created and presented using the language GNU Octave \protect\cite{eaton:2002} and the Psychophysics Toolbox extensions \protect \cite{ptbx1, ptbx2}.
%}
%
%\subsubsection{\deleted{Design}}
%\deleted{
%We presented participants with 64 trials in experiment one in a single block. The basic properties of a trial were that it consisted of two elements: (1) two squares each containing a diffferent number of dots, and (2) a referring expression that referred to one of the squares by indicating the number of dots that it contained, leaving the other as a distractor. We measured two dependent variables: response time, and response accuracy. It was always the case that two squares appeared on the screen, each containing a number of dots: therefore items can be described using the notation $pair_{(p=1... 8)}([n],m)$ to indicate that for the stimulus pair $p$, one of the numbers of dots was $n$ and the other $m$; and that the square-bracketed element $[n]$ was the target. There were 8 such pairs : $pair_{(1)}(2,4)$; $pair_{(2)}(3,5)$; $pair_{(3)}(2,6)$; $pair_{(4)}(3,7)$; $pair_{(5)}(6,8)$; $pair_{(6)}(7,9)$; $pair_{(7)}(4,8)$; $pair_{(8)}(5,9)$. See Table \ref{tablee1} for the properties of the stimuli.
%}
%
%
%\begin{table}[tbp]
%\caption{\deleted{Properties of stimuli used in Experiment 1, arranged by pair}}
%\label{tablee1}
%\begin{tabular}{lll}
%\deleted{subitizabiity}	& \deleted{gap size}  	&	\deleted{contents}\\
%\hline
%\deleted{subitizable}	&\deleted{small gap}	&	\deleted{$pair_{(1)}(2,4)$}\\
% 			&			&	\deleted{$pair_{(2)}(3,5)$}\\
%\hline

>>>>

\documentclass[doc,apacite]{apa6}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{color}
\usepackage{booktabs}
\usepackage{layout}
\usepackage{apacite}
\usepackage{datetime}
\usepackage{rotating}
\title{Vagueness as cost reduction}
\date{\today}
\authornote{\today ~\currenttime}
\shorttitle{Vagueness as cost reduction}
\twoauthors{Matt Green}{Kees van Deemter}
\twoaffiliations{Computing Science\\ University of Aberdeen}{Computing Science\\ University of Aberdeen}

\abstract{Much of everyday language is vague, yet standard game-theoretic models make it difficult to see how vague expressions can have benefits over crisp ones in co-operative situations. We report a series of experiments that aims to separate the utility of vagueness from the utility of other factors that come together with vagueness. We argue that the results support a view of vagueness where the benefits that vague terms exert are due to other influences that vagueness brings with it rather than to vagueness itself.}

@@

Linguists, philosophers of language, and more recently game theorists, have asked why natural languages contain so many vague expressions, which are used so frequently \cite{Lipman:2000fk, lipmanvague}. By introducing borderline cases, these expressions create potential misunderstandings, thereby creating ``a worldwide several-thousand year efficiency loss'' \cite<>[p.~1]{lipmanvague}. Lipman explains the point by means of a scenario in which a speaker describes a person to a hearer, who needs to identify that person in the arrivals hall of of an airport. Lipman argues that, in such a scenario, a precise description of the person's height (e.g., ``The person's height is 187.96 cm'') would be more useful than a vague one (``The person is tall''). Lipman uses this scenario to explain why standard game theory models of communication \cite<e.g.,>{Crawford:1982lr} predict that, under certain conditions, a crisp act of communication will always have more utility than a vague act that communicates the same state of affairs. The relevant conditions are, broadly speaking, that both interlocutors know all the relevant facts (e.g., both know the person's height precisely) and that the setting for the communication is co-operative. These conditions exclude deliberate deception, and rhetorical situations like political debate and advertising where the intention is to persuade the interlocutor to adopt some point of view, where the persuasion might not be intended to be in the addressee's best interest.

Lipman argued that such the efficiency loss resulting from vague expressions would be unlikely to have arisen unless there are advantages as well as disadvantages associated with vague expressions. Lipman asked, essentially, what these advantages might be, and how they might find a place in a game-theoretical explanation. In this paper, we focus on the first part of Lipman's question.

Several tentative answers to Lipman's question have been offered  \cite<see>{van2009utility, van2010vagueness}. Prominent among these answers is the idea that vague expressions are somehow easier to process, by a speaker and/or a hearer, than expressions that are not vague (i.e., crisp) \cite<e.g.,>[]{lipmanvague,De-Jaegher:2003lr,vanrooij2003lr}. For example, \citeA<>[p.\ 11]{lipmanvague} writes: ``For the listener, information which is too specific may require more effort to analyze''. We shall refer to this characterisation of the utility of vague language as the \emph{cost reduction} hypothesis. The idea of vagueness as cost reduction can take various shapes, but the basic hypothesis is that it is easier for people to think in terms of loosely defined categories (such as ``quite a few'', or ``many'') than in terms of crisply defined ones (such as ``thirteen'', or ``237''). This predicts that whenever a vague expression can perform the same communicative task as a precise one, it is rational to choose the vague expression. The corollary for comprehension is that vague expressions should be understood more readily than precise expressions.

Questions concerning optimal language use have many practical applications. Natural Language Generation\footnote{NLG systems take data or formulas as input, and transform them into natural language outputs \cite<>[]{reiter2000building}. The process parallels language production in humans.}  (henceforth NLG) systems must make decisions between different formulations of the same information. For example, if a man's height is 6 foot 2 inches, this could be expressed as ``187.96 metres'', ``6 foot 2'', or ``tall'', among other ways, and the NLG system must decide between these.  The problem is particularly relevant for NLG systems that take numbers as input, as many do. In the context of an NLG system faced with a practical decision of this kind, Lipman's question becomes ``Under what circumstances should vague terms be produced?''  Relevant applications include weather forecasting on the basis of numerical weather data such as temperature and wind speed \cite{goldberg1994using,turner2006generating}, and medical decision support on the basis of clinical measurement such as oxygen saturation, heart rhythm, etc. \cite{Hripcsak01032009, hunter2008summarising, portet2009automatic}. At present, such NLG systems are often forced to make decisions concerning the level of precision in the utterances that they generate (e.g., ``the temperature will be in the high twenties tomorrow'') on the basis of little more than intuition. A better understanding of the benefits of different precision levels for readers would allow these systems to become more useful.

@@

In a series of studies of behaviour modification, \citeA<>{Mishra01042011} manipulated the presentation format of information about quantities in the domains of mental acuity, physical strength, and weight loss. In the weight loss study, participants were told that the study was designed to test the validity of a new (actually fictitious) health index, the HHI (Holistic Health Index). They were told that an ideal HHI score lies in the range of 45 to 55. In a longitudinal study, participants submitted their height, weight, hydration level, gender, and age to a computer each week. Participants were told that two algorithms would be used to compute their HHI, and that it was possible that the two algorithms might give different values initially, but would converge over the course of the study to a single value. They were also told that if the two algorithms did give different values, then the true score lay between the two values. In one condition, which the authors called the precise condition, the two algorithms gave the same score. In the other condition, which the authors called the vague condition, one algorithm added 3\% to the score while the other algorithm subtracted 3\% from the score, yielding a range of values whose midpoint was the same as the two values given in the precise condition.

One group of participants was given HHI scores in the ideal range: for this group their weight loss did not differ depending on whether they were given vague or precise HHI values. However for the other group, who were given HHI scores outside the ideal range, their weight loss was significantly greater if they were given vague HHI scores than if they were given precise HHI scores. The authors explain the improvement in the vague condition for this group as resulting from the participants' freedom to think of themselves a positioned on one end of the range - the end closest to the ideal HHI scores. This ``illusion of proximity'' \cite<>[p.~4]{Mishra01042011} to the goal is argued to allow participants to generate positive expectancies that lead to behaviours that improve performance. In contrast, in the precise conditions, participants did not have this freedom of interpretation, and could not distort the information to bring about the beneficial \emph{illusion of proximity}. These results are interesting, and of obvious potential practical importance. We note, however, that information presented as an exact range of values does not conform with the standard definition of vagueness \cite{keefe1996vagueness, EgreKlinedinst}, since an exact range does not admit borderline cases. In the terminology of \citeA{Hobbs85granularity}, the difference between a range and a single midpoint value is a difference of \emph{granularity}. Furthermore, the experiments of \citeA{Mishra01042011} did not explore benefits in terms of processing cost, but in terms of long-term behaviour change.

Similar issues arise from the work of \citeA{peters2009bringing}. The authors carried out a series of studies where participants were required to rate hospitals based on various sources of information about quality of care. There was a between-subjects manipulation based on numeracy. The format of the information was manipulated within subjects: either numbers only were presented, or both numbers and evaluative categories were presented (e.g., \emph{Poor}, \emph{Fair}, \emph{Good}, \emph{Excellent}, with crisp visual boundary lines between the categories). Results showed that, for low-numeracy participants, the presence of evaluative categories resulted in a diminished influence of an irrelevant affective state on the ratings. For all participants, the presence of evaluative categories resulted in better decisions and in a greater use of the most important and reliable types of information, such as survival rates.

It is, however, questionable whether the ``evaluative categories" manipulation in this study can be considered a manipulation of vagueness. Certainly, terms like \emph{Fair} admit the possibility of borderline cases. However, given that the boundaries between the categories were marked crisply, and that therefore the categories mapped crisply to numerical values, it becomes doubtful whether any borderline cases could be conceived to arise in fact. For example, \emph{Fair} was mapped to 60\% -- 70\% for the variable \emph{percentage of heart attack patients given recommended treatment (ACE inhibitor)}. Accordingly, rather than the vagueness of categories such as \emph{Poor}, Peters et al. emphasise the evaluative content inherent in these categories, and the affective potential of the evaluative content rather than the vagueness of the terms like {\em Fair}.
%\\[2ex]
\subsection{}
The experiments reported in the present paper put the cost reduction hypothesis to the test. The question that we are trying to answer is whether vague expressions are processed more easily by readers than crisp ones. Like Lipman, we focus on situations where numerical information is used in order to identify a referent. Reference, in other words, will be the linguistic task on which we focus, partly because of the interest that this topic has recently drawn from the NLG community.  In focussing on benefits for the hearer, we will leave aside the question of audience design, leaving this for later research.

In using references to quantities to test the cost reduction hypothesis we are only testing one aspect of vagueness in a particular context. This limits the applicability of our results. It also has the advantage that we can explore the costs and benefits of vagueness more thoroughly in that context. Since one prevalent view of vagueness is that a vague expression is never preferable to a crisp equivalent, a demonstration of a benefit for vagueness in any context would advance the discussion.

In our experiments we used a speeded forced choice task to compare the processing costs of different references to quantities. In this context, speed and accuracy of responses are the key dimensions on which the different references can be compared. The stimuli in the experiments were sets of dot arrays containing various numbers of dots. The forced choice was to identify one dot array given a reference to a given quantity of dots. We manipulated the references in several ways across a series of four experiments.

Our main manipulation was always of vagueness: we constructed crisp and vague versions of references to the same dot array. For example, in experiment 1 the instruction presented to the participant could be \emph{Choose the square with many dots.} (vague condition), or \emph{Choose the square with 20 dots.} (crisp condition), identifying the same dot array.

Ideally, we would manipulate vagueness independently of other variables. However, in practice, and given the constraints imposed by the use of natural language, any manipulation of vagueness introduces variance along other dimensions as well. Therefore we carried out further experiments to try to address and control variance along other dimensions.

One non-vagueness source of variance was the difference between numerical and verbal format in the references. For example, in the instructions for experiment 1 given above, \emph{many} is in verbal format (as well as vague) and \emph{20} is in numerical format (as well as crisp). Therefore any difference observed between responses to the \emph{many} instruction and the \emph{20} instruction could be due either to vagueness or to {\bf instruction format}. It is possible to create vague and crisp references in each of these instruction formats: for example in numerical instruction format we can have \emph{20} (crisp) and \emph{about 20} (vague), and in verbal instruction format we can have \emph{the most} (crisp) and \emph{many} (vague). Experiment 2 varied vagueness and instruction format factorially in a similar way to this, to try to tease apart these two sources of variance.

Another potential non-vagueness source of variance between responses to the \emph{many} (verbal vague) references and the \emph{20} (numeric crisp) references lies in the method used to identify a referent (we call this the {\bf selection algorithm} source of variance). Identifying the array with \emph{many} dots might be done by a \emph{comparison} algorithm, identifying that one array is more numerous than the others without establishing the numerosity of any arrays, whereas identifying the array with \emph{20} dots, which might be done by a \emph{matching} algorithm would require an estimate of the numerosity of the arrays. It seems reasonable that comparison would be faster than matching because it does not require estimates of cardinality. Furthermore the problem persists when using references like \emph{the most} (verbal crisp) and \emph{about 20} (numeric vague), since they too vary not only along the vague/crisp dimension and the numeric/verbal dimension but also along the comparison/matching dimension. Experiments 3 and 4 addressed the \emph{selection algorithm} source of variance - experiment 3 crossed vague/crisp and comparison/matching factorially using only numerical references, and experiment 4 crossed vague/crisp and comparison/matching factorially using only verbal references.

\section{Experiment One}

\subsection{Introduction}
We used a forced choice task to compare choices made in response to vague instructions against choices made in response to crisp instructions. The participant was presented with an instruction like ``Choose the square with many dots'' in the vague conditions, or ``Choose the square with 20 dots'' in the crisp conditions. Then two dot arrays were presented in the form of squares containing a number of dots. Fig \ref{stimuluse1} shows an example stimulus. The participant was required to identify the square that corresponded with the instruction, by pressing the appropriate key.
Response time and accuracy were recorded for analysis.

Our main manipulation was of the vagueness of the instruction, with two levels, vague and crisp. Table \ref{instructionse1} shows examples from each condition.  We also manipulated how discriminable the dot arrays were. One array always contained 25 dots: the other contained either 5, 10, 15, 20, 30, 35, 40, or 45 dots. This led to numerical differences of 5, 10, 15, and 20, with lower differences resulting in less discriminable arrays and larger differences representing more discriminable arrays. There is evidence that when the distance grows between two numbers, they become more easily distinguishable from each other: the \emph{numerical distance effect}, which has been shown for comparing the numerosity of two sets of dots \cite{van123} and for processing Arabic numerals and number words \cite{Dehaene199647}. Where a number was mentioned in the instruction, it was always in the form of an Arabic numeral.

The instructions indicated the larger of the two dot arrays equally often as they indicated the smaller of the dot arrays, so that participants could not systematically choose the larger or smaller array as a successful response strategy. There is evidence that when two numbers are presented with the smaller on the left, this left-side presentation facilitates responses indicating the smaller number: the \emph{Spatial-Numerical Association of Response Codes (SNARC)} effect \cite{dehaene1993mental, gevers2006automatic}. We controlled which side the smaller number appeared on to avoid systematic influences of this effect.

\begin{table}[htbp]
\caption{Table of instructions for the pair (5,25). Experiment 1}
\label{instructionse1}
\begin{tabular}{rl}
\toprule
vagueness&example\\
\midrule
crisp 	& 	Choose the square with 5 dots \\
vague	&	Choose the square with few dots\\
\bottomrule
\end{tabular}
\end{table}

\subsection{Hypotheses}
The {\bf cost reduction} hypothesis predicts that responses will be faster and more accurate for the vague conditions than for the crisp conditions because the vague conditions impose a lower cognitive load than the crisp conditions.

The {\bf instruction format} hypothesis predicts that responses will be faster and more accurate in the vague (i.e., verbal) conditions than in the crisp (i.e., numerical) conditions because numbers are harder to process than verbal references to quantities.

The {\bf selection algorithm} account predicts that responses will be faster and more accurate in the vague (i.e., comparison) conditions than in the crisp (i.e., matching) conditions because it is easier to carry out comparison than matching.

Thus all three accounts predict a main effect advantage of vagueness, but for different reasons. All three accounts also predict a main effect advantage for greater numerical distance, because greater numerical distances lead to more easily discriminable arrays. The accounts also share the prediction of an interaction effect whereby the main effect advantage of vagueness should diminish as numerical distance increases.
